\setNextFileName{GetTheSource.html}
\begin{section}{Get the Source}
\label{sec:getthesource}

The source code for the Jikes RVM is stored in a \href{http://mercurial.selenic.com/}{Mercurial} repository. You can browse the online mercurial repository at \url{http://hg.code.sourceforge.net/p/jikesrvm/code}.

A developer can either work with the version control system or download one of the releases. If you are interested in doing development of Jikes RVM you should probably use Mercurial instead of downloading a release.

\begin{subsection}{Download a Release}

Major and minor releases of Jikes RVM occur at regular intervals. These releases are archived in the \href{http://sourceforge.net/projects/jikesrvm/files/}{file download area} in either tar-gzip (jikesrvm-<version>.tar.gz) or tar-bzip2 (jikesrvm-<version>.tar.bz2) format. Use your web browser to download the latest version of Jikes RVM then to extract the tar-gzip archive type:

\begin{lstlisting}
$ tar xvzf jikesrvm-<version>.tar.gz
\end{lstlisting}

or for the tar-bzip2 archive type:

\begin{lstlisting}
$ tar xvjf jikesrvm-<version>.tar.bz2
\end{lstlisting}

\end{subsection}

\begin{subsection}{Use Mercurial}

The source code for Jikes RVM is stored in a Mercurial repository. Mercurial and other distributed revision control systems (e.g. Git) are quite different from centralized version control systems like CVS and Subversion. If you are not familiar with Mercurial, you can find instructions on Mercurial use at \url{http://mercurial.selenic.com/guide/}. There is also a \href{http://hgbook.red-bean.com/}{Mercurial book}.

After installing Mercurial the current version of source can be downloaded via:

\begin{lstlisting}
$ hg clone http://hg.code.sourceforge.net/p/jikesrvm/code
\end{lstlisting}

This will clone the Jikes RVM repository into the newly created directory jikesrvm.

If you need a specific version, it is recommended to clone the complete repository nonetheless. You can then switch to a specific release, e.g. 2.4.6, by doing the following:

\begin{lstlisting}
$ cd jikesrvm
$ hg checkout 2.4.6
\end{lstlisting}

If you are a not core developer you will not be able to push changes to the main Jikes RVM repository directly. If you want to contribute to the Jikes RVM, please take a look at this \href{http://www.jikesrvm.org/Contributions/}{page}.

\end{subsection}

\end{section}
