\setNextFileName{UnboxedTypes.html}
\begin{section}{Unboxed Types}
\label{sec:unboxedtypes}

If a type is boxed then it means that values of that type are represented by a pointer to a heap object. An unboxed type is represented by the value itself such as \spverb+int+, \spverb+double+, \spverb+float+, \spverb+byte+ etc.

In the Jikes RVM terminology, an unboxed type is a custom unboxed type. Normal Java primitives such as \spverb+int+ are never referred to as unboxed types.

The Jikes RVM also defines a number of unboxed types. Due to a limitation of the way the compiler generates code the Jikes RVM must define an unboxed array type for each unboxed type. The unboxed types are:
\begin{itemize}
  \item \spverb+org.vmmagic.unboxed.Address+
  \item \spverb+org.vmmagic.unboxed.Extent+
  \item \spverb+org.vmmagic.unboxed.ObjectReference+
  \item \spverb+org.vmmagic.unboxed.Offset+
  \item \spverb+org.vmmagic.unboxed.Word+
  \item \spverb+org.jikesrvm.compilers.common.Code+
\end{itemize}

Values of unboxed types appear only in the virtual machine's stack, registers, or as fields/elements of class/array instances.

Unboxed types may inherit from Object but they are not objects. As such there are some restrictions on the use of unboxed types:
\begin{itemize}
  \item A unboxed type instance must not be passed where an \spverb+Object+ is expected. This will type-check, but it is not what you want. A corollary is to avoid overloading a method where the two overloaded versions of the method can only be distinguished by operating on an Object versus an unboxed type. The optimizing compiler can detect \textit{some} invalid uses of unboxed types.
  \item An unboxed type must not be synchronized on.
  \item They have no virtual methods.
  \item They do not support lock operations, generating hashcodes or any other method inherited from \spverb+Object+.
  \item All methods must be compiler intrinsics.
  \item  Avoid making an array of an unboxed type. Instead represent it by the array version of unboxed type. i.e. \spverb+org.vmmagic.unboxed.Address[]+ should be replaced with \texttt{org.vm\-ma\-gic.un\-box\-ed.Address\-Ar\-ray} but \texttt{org.vm\-ma\-gic.un\-box\-ed.Address\-Ar\-ray[]} is fine.
\end{itemize}


\end{section}
